---
title: "Vehicle Analysis"
author: "Glorija"
date: "2024-07-02"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(tidyr)
library(caret)
library(ranger)
library(ggplot2)
library(gridExtra)
library(mgcv)
library(progressr)
library(car)
library(glmnet)
library(Hmisc)


data <- read.csv("C:\\Users\\glori\\Desktop\\R studies\\Vehicle Analysis\\clean_vehicle_data.csv")

```

## Introduction

This is a presentation of my findings while investigating the dataset car_prices

### The dataset has been cleaned and filtered:

- Removed irrelevant columns such as: trim, vin, body, state, interior, body, seller, and sell date;
- Cleared non-sensical data, missing values, and NAs;
- Made unique makes and colors uniform;

## Data Summary

```{r Summary, echo = TRUE}
summary(data)
head(data)
str(data)
```

## EDA
- I used a correlation matrix to check numerical predictor relationship with the target and Linear Regression to assess the F and p values, as well as ANOVA test for categorical variables

```{r eda}
## Correlation Analysis for Numerical Variables
# Numerical Variables - Correlation Matrix
numerical_vars <- data %>% select(year, mmr, odometer, condition, sellingprice)
cor_matrix <- cor(numerical_vars)
high_corr <- findCorrelation(cor_matrix, cutoff = 0.75)
selected_numerical_vars <- numerical_vars[ , -high_corr]

# Display correlation matrix and selected numerical variables
cat("### Correlation Matrix:\n")
print(cor_matrix)

# ANOVA
anova_full <- aov(sellingprice ~ make + color + transmission, data = data)
cat("### ANOVA:\n")
summary(anova_full)

# Perform linear regression
cat("###Linear Regression:\n")
model_numerical <- lm(sellingprice ~ year + mmr + odometer + condition, data = data)
cat("### Linear Regression:\n")
summary(model_numerical)
```

## Random Forest Model
- Correlation matrix and ANOVA didn't return desirable output so I used Random Forest for its robustness and ability to handle complex interactions between variables without requiring much preprocessing (for data with numerical and categorical variables such as this one)
- Random Forest accounts for interaction between variables rather just individual relationships between target and predictors, unlike in filter methods
- Random Forests are robust to overfitting, especially when dealing with large datasets, because they average the predictions of multiple trees, reducing variance
- Random Forest can model both linear and non-linear relationships by building multiple decision trees that split the data based on different predictor values. Each tree captures different aspects of the data, including linear and non-linear patterns
- I also displayed the importance of each predictor in predicting the target variable, and used Random Forest model as a variable selection method

```{r random forest}
data <- read.csv("C:\\Users\\glori\\Desktop\\R studies\\Vehicle Analysis\\clean_vehicle_data.csv")

# Convert condition to numeric for modeling
data$condition <- as.numeric(as.character(data$condition))

# Select predictors and target variable
predictors <- data[, c("year", "mmr", "odometer", "condition", "make", "color", "transmission")]
target <- data$sellingprice

# Train a Random Forest model
rf_model <- ranger(
  formula = sellingprice ~ .,
  data = data,
  num.trees = 100,
  importance = 'impurity',
  verbose = TRUE
)

# Print variable importance
importance_scores <- importance(rf_model)
importance_df <- data.frame(Variable = names(importance_scores), Importance = importance_scores)

# Convert importance to percentages
importance_df$Importance <- importance_df$Importance / sum(importance_df$Importance) * 100

# Plot variable importance with percentages
ggplot(importance_df, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(Importance, 2), "%")), hjust = -0.2, size = 3.5) +
  coord_flip() +
  labs(title = "Variable Importance from Random Forest Model", x = "Variable", y = "Importance (%)") +
  theme_minimal()

# Predict on the training data
rf_predictions <- predict(rf_model, data = data)$predictions

# Calculate R-squared and other metrics
rf_r_squared <- cor(rf_predictions, data$sellingprice)^2
rf_rmse <- sqrt(mean((data$sellingprice - rf_predictions)^2))
rf_mae <- mean(abs(data$sellingprice - rf_predictions))

# Print the metrics
print(paste("Random Forest R-squared:", rf_r_squared))
print(paste("Random Forest RMSE:", rf_rmse))
print(paste("Random Forest MAE:", rf_mae))

# Scatter plots to explore relationships between numerical features and selling price
p1 <- ggplot(data, aes(x = year, y = sellingprice)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", col = "red") + 
  ggtitle("Year vs Selling Price")

p2 <- ggplot(data, aes(x = mmr, y = sellingprice)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", col = "red") + 
  ggtitle("MMR vs Selling Price")

p3 <- ggplot(data, aes(x = odometer, y = sellingprice)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", col = "red") + 
  ggtitle("Odometer vs Selling Price")

p4 <- ggplot(data, aes(x = as.numeric(condition), y = sellingprice)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", col = "red") + 
  ggtitle("Condition vs Selling Price")

# Arrange plots
grid.arrange(p1, p2, p3, p4, ncol = 2)

## VIF
# Assuming selected_vars is the vector containing selected variable names
selected_vars <- c("year", "mmr", "odometer", "condition")

# Construct the formula
formula <- as.formula(paste("sellingprice ~", paste(selected_vars, collapse = " + ")))

# Fit the model
model <- lm(formula, data = data)
vif_values <- vif(model)
print(vif_values)
```

- I used Scatter plots to visualize the relationship between numerical features (year, mmr, odometer, condition) and the selling price
-I added a linear regression line to show the trend between the variables

## GAM
- Since 3 of the 4 predictors had non-linear relationships with the target,
I chose GAM model for its flexibility in capturing the non-linear relationships
- GAMs extend linear models by allowing non-linear functions of predictors. They use smooth functions (splines) to model the relationship between each predictor and the target variable, accommodating non-linearity
- GAMs fit each predictor’s effect separately, which can be advantageous when different predictors have different types of non-linear relationships with the target variable
```{r gam}
# Fit a GAM model
gam_model <- gam(sellingprice ~ s(year) + s(mmr) + s(odometer) + s(condition), data = data)

# Summary of GAM model
summary(gam_model)

# Plot smooth terms
par(mfrow = c(2, 2))
plot(gam_model, shade = TRUE, pages = 1)

# Predict on the training data
gam_predictions <- predict(gam_model, newdata = data)

# Calculate R-squared and other metrics
gam_r_squared <- cor(gam_predictions, data$sellingprice)^2
gam_rmse <- sqrt(mean((data$sellingprice - gam_predictions)^2))
gam_mae <- mean(abs(data$sellingprice - gam_predictions))

# Print the metrics
print(paste("GAM R-squared:", gam_r_squared))
print(paste("GAM RMSE:", gam_rmse))
print(paste("GAM MAE:", gam_mae))
```

## Random Forest Cross-Validation
- I used k-Fold Cross-Validation to assess the model's performance by dividing the data into k subsets, training on k-1 subsets, and testing on the remaining subset
- Example usage with different k values (k_values <- c(5, 10, 15)), has proven that 10-Fold Cross-Validation provided a good balance between bias and variance in the performance estimate

```{r rf cross-validation}
# Define the formula for the Random Forest model
formula_selected <- sellingprice ~ year + mmr + odometer + condition
# Define k-fold cross-validation function for Random Forest with additional metrics
cross_validate_rf <- function(data, formula, k_folds = 10, num_trees = 100) {
  folds <- createFolds(data$sellingprice, k = k_folds, list = TRUE, returnTrain = TRUE)
  rmse_values <- numeric(k_folds)
  r_squared_values <- numeric(k_folds)
  mae_values <- numeric(k_folds)
  
  for (i in 1:k_folds) {
    train_indices <- folds[[i]]
    train_data <- data[train_indices, ]
    test_data <- data[-train_indices, ]
    rf_model <- ranger(formula, data = train_data, num.trees = num_trees, importance = 'impurity')
    predictions <- predict(rf_model, data = test_data)$predictions
    rmse_values[i] <- sqrt(mean((test_data$sellingprice - predictions)^2))
    r_squared_values[i] <- cor(predictions, test_data$sellingprice)^2
    mae_values[i] <- mean(abs(test_data$sellingprice - predictions))
  }
  
  mean_rmse <- mean(rmse_values)
  mean_r_squared <- mean(r_squared_values)
  mean_mae <- mean(mae_values)
  
  return(list(rmse = mean_rmse, r_squared = mean_r_squared, mae = mean_mae))
}

# Perform cross-validation
rf_cv_results <- cross_validate_rf(data, formula_selected, k_folds = 10, num_trees = 100)
print(paste("Average RMSE from k-fold cross-validation (RF):", rf_cv_results$rmse))
print(paste("Average R-squared from k-fold cross-validation (RF):", rf_cv_results$r_squared))
print(paste("Average MAE from k-fold cross-validation (RF):", rf_cv_results$mae))

```

## GAM Cross-Validation

```{r gam cross-validation}

formula_gam <- sellingprice ~ s(year) + s(mmr) + s(odometer) + s(condition)
# Define k-fold cross-validation function for GAM with additional metrics
cross_validate_gam <- function(data, formula, k_folds = 10) {
  folds <- createFolds(data$sellingprice, k = k_folds, list = TRUE, returnTrain = TRUE)
  rmse_values <- numeric(k_folds)
  r_squared_values <- numeric(k_folds)
  mae_values <- numeric(k_folds)
  
  for (i in 1:k_folds) {
    train_indices <- folds[[i]]
    train_data <- data[train_indices, ]
    test_data <- data[-train_indices, ]
    gam_model <- gam(formula, data = train_data)
    predictions <- predict(gam_model, newdata = test_data)
    rmse_values[i] <- sqrt(mean((test_data$sellingprice - predictions)^2))
    r_squared_values[i] <- cor(predictions, test_data$sellingprice)^2
    mae_values[i] <- mean(abs(test_data$sellingprice - predictions))
  }
  
  mean_rmse <- mean(rmse_values)
  mean_r_squared <- mean(r_squared_values)
  mean_mae <- mean(mae_values)
  
  return(list(rmse = mean_rmse, r_squared = mean_r_squared, mae = mean_mae))
}

# Perform cross-validation for GAM
gam_cv_results <- cross_validate_gam(data, formula_gam, k_folds = 10)
print(paste("Average RMSE from k-fold cross-validation (GAM):", gam_cv_results$rmse))
print(paste("Average R-squared from k-fold cross-validation (GAM):", gam_cv_results$r_squared))
print(paste("Average MAE from k-fold cross-validation (GAM):", gam_cv_results$mae))

```

## Conlusion

```{r conclusion}
cat("\nModel Selection Summary:\n")
cat("1. Both GAM and Random Forest models show 
    comparable performance in predicting vehicle selling prices.\n")
cat("2. The GAM model has a slightly lower RMSE (", gam_cv_results$rmse, ") 
    compared to the Random Forest model (", rf_cv_results$rmse, ").\n")
cat("3. The R-squared for GAM (", gam_cv_results$r_squared, ") is also 
    slightly higher than for Random Forest (", rf_cv_results$r_squared, ").\n")
cat("4. The MAE for GAM (", gam_cv_results$mae, ") is 
    comparable to that for Random Forest (", rf_cv_results$mae, ").\n")
cat("5. Given the small difference in RMSE and R-squared, and 
    considering factors like interpretability and computational 
    efficiency, the GAM model is preferred for its simplicity 
    and better performance.\n")

```

## Model Fitting

```{r model fitting}
data <- read.csv("C:\\Users\\glori\\Desktop\\R studies\\Vehicle Analysis\\clean_vehicle_data.csv")

# Define the formula for the final GAM model
formula_final <- sellingprice ~ s(year) + s(mmr) + s(odometer) + s(condition)

# Fit the GAM model to the full dataset
final_gam_model <- gam(formula_final, data = data)

# Check the summary of the final GAM model
summary(final_gam_model)

plot(final_gam_model, pages = 1)
plot(final_gam_model$residuals)

# Residuals vs Fitted
plot(final_gam_model$fitted.values, final_gam_model$residuals, xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red")

# Calculate RMSE on the full data
predictions <- predict(final_gam_model, newdata = data)
rmse <- sqrt(mean((data$sellingprice - predictions)^2))
print(paste("RMSE on the full dataset:", round(rmse, 2)))
```
